---
layout: post
title:  Data Processing with dplyr & tidyr
date: 2015-09-07
published: true
tags: [r, data wrangling, dplyr, tidyr]
---


Analysts tend to follow 4 fundamental processes to turn data into understanding, knowledge & insight:

1. **Data manipulation**
2. Data visualization
3. Statistical analysis/modeling
4. Deployment of results

This tutorial will focus on **<u>data manipulation</u>**

<br><br>

### <u>Data Manipulation</u>

> It is often said that 80% of data analysis is spent on the process of cleaning and preparing the data. (Dasu and Johnson, 2003)

Well structured data serves two purposes:

1. Makes data suitable for software processing whether that be mathematical functions, visualization, etc.
2. Reveals information and insights

Hadley Wickham's paper on [Tidy Data](http://vita.had.co.nz/papers/tidy-data.html) provides a great explanation behind the concept of "tidy data"


<br>

### <u>Why Use tidyr & dplyr</u>
- Although many fundamental data processing functions exist in R, they have been a bit convoluted to date and have lacked consistent coding and the ability to easily *flow* together &#8594; leads to difficult-to-read nested functions and/or *choppy* code.
- [R Studio](http://www.rstudio.com/) is driving a lot of new packages to collate data management tasks and better integrate them with other analysis activities &#8594; led by [Hadley Wickham](https://twitter.com/hadleywickham) & the R Studio [team](http://www.rstudio.com/about/) &#8594; [Garrett Grolemund](https://twitter.com/StatGarrett), [Winston Chang](https://twitter.com/winston_chang), [Yihui Xie](https://twitter.com/xieyihui) among others.
- As a result, a lot of data processing tasks are becoming packaged in more cohesive and consistent ways &#8594; leads to:
    - More efficient code
    - Easier to remember syntax
    - Easier to read syntax

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```

> <em>tidyr and dplyr packages provide fundamental functions for cleaning, processing, & manipulating data</em>

* tidyr
    + <a href="#gather">`gather()`</a>
    + <a href="#spread">`spread()`</a>
    + <a href="#separate">`separate()`</a>
    + <a href="#unite">`unite()`</a>
* dplyr
    + <a href="#select">`select()`</a>
    + <a href="#filter">`filter()`</a>
    + <a href="#group">`group_by()`</a>
    + <a href="#summarise">`summarise()`</a>
    + <a href="#arrange">`arrange()`</a>
    + <a href="#join">`join()`</a>
    + <a href="#mutate">`mutate()`</a>

<br>

<a href="#">Go to top</a>

<center><hr width="50%"></center>

<br>

## <u>%>% Operator</u>
Although not required, the tidyr and dplyr packages make use of the pipe operator `%>%` developed by [Stefan Milton Bache](https://twitter.com/stefanbache) in the R package [magrittr](http://cran.r-project.org/web/packages/magrittr/magrittr.pdf).  Although all the functions in tidyr and dplyr *can be used without the pipe operator*, one of the great conveniences these packages provide is the ability to string multiple functions together by incorporating `%>%`.

This operator will forward a value, or the result of an expression, into the next function call/expression.  For instance a function to filter data can be written as:

<center>filter(data, variable == <em>numeric_value</em>)</center>

<center><em><u>or</u></em></center>

<center>data %>% filter(variable == <em>numeric_value</em>)</center>

<br>

Both functions complete the same task and the benefit of using `%>%` is not evident; however, when you desire to perform multiple functions its advantage becomes obvious.  For instance, if we want to filter some data, summarize it, and then order the summarized results we would write it out as:

&nbsp;&nbsp;<u>Nested Option:</u>

&nbsp;&nbsp;&nbsp;&nbsp;arrange(<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;summarize(<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;filter(data, variable == *numeric_value*),<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Total = sum(variable)<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;),<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;desc(Total)<br>
&nbsp;&nbsp;&nbsp;&nbsp;)

<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;*<u>or</u>*
<br>

&nbsp;&nbsp;<u>Multiple Object Option:</u>

&nbsp;&nbsp;&nbsp;&nbsp;  a <- filter(data, variable == *numeric_value*)<br>
&nbsp;&nbsp;&nbsp;&nbsp;  b <- summarise(a, Total = sum(variable))<br>
&nbsp;&nbsp;&nbsp;&nbsp;  c <- arrange(b, desc(Total))<br>

<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;*<u>or</u>*
<br>

&nbsp;&nbsp;<u>%>% Option:</u>

&nbsp;&nbsp;&nbsp;&nbsp; data %>%<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;filter(variable == "value") %>%<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;summarise(Total = sum(variable)) %>%<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;arrange(desc(Total))

<br>

As your function tasks get longer the `%>%` operator becomes more efficient *<u>and</u>* makes your code more legible.  In addition, although not covered in this tutorial, the `%>%` operator allows you to flow from data manipulation tasks straight into vizualization functions *(via ggplot and ggvis)* and also into many analytic functions.

To learn more about the `%>%` operator and the magrittr package visit any of the following:

- [http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)
- [http://www.r-bloggers.com/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/](http://www.r-bloggers.com/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/)
- [http://blog.revolutionanalytics.com/2014/07/magrittr-simplifying-r-code-with-pipes.html](http://blog.revolutionanalytics.com/2014/07/magrittr-simplifying-r-code-with-pipes.html)

<br>

<a href="#">Go to top</a>

<center><hr width="50%"></center>

<br>

## <u>tidyr Operations</u>

There are four fundamental functions of data tidying:

* <a href="#gather">`gather()`</a> takes multiple columns, and gathers them into key-value pairs: it makes “wide” data longer
* <a href="#spread">`spread()`</a> takes two columns (key & value) and spreads in to multiple columns, it makes “long” data wider
* <a href="#separate">`separate()`</a> splits a single column into multiple columns
* <a href="#unite">`unite()`</a> combines multiple columns into a single column

<br><br>

<a id="gather"> </a>

### <u>gather( ) function</u>: 
**Objective:** Reshaping wide format to long format

**Description:** There are times when our data is considered unstacked and a common attribute of concern is spread out across columns.  To reformat the data such that these common attributes are *gathered* together as a single variable, the `gather()` function will take multiple columns and collapse them into key-value pairs, duplicating all other columns as needed.

**Complement to:**  <a href="#spread">`spread()`</a>


```{r, echo=TRUE, eval=FALSE}
Function:       gather(data, key, value, ..., na.rm = FALSE, convert = FALSE)
Same as:        data %>% gather(key, value, ..., na.rm = FALSE, convert = FALSE)

Arguments:
        data:           data frame
        key:            column name representing new variable
        value:          column name representing variable values
        ...:            names of columns to gather (or not gather)
        na.rm:          option to remove observations with missing values (represented by NAs)
        convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or 
                        factor as appropriate
```

<br>

**Example**

We'll start with the following data set:
```{r, echo=FALSE, message=FALSE}
DF <- tbl_df(read.table(header = TRUE, text = "
   Group   Year   Qtr.1  Qtr.2  Qtr.3  Qtr.4
    1   2006    15      16      19    17
    1   2007    12      13      27    23
    1   2008    22      22      24    20
    1   2009    10      14      20    16
    2   2006    12      13      25    18
    2   2007    16      14      21    19
    2   2008    13      11      29    15
    2   2009    23      20      26    20
    3   2006    11      12      22    16
    3   2007    13      11      27    21
    3   2008    17      12      23    19
    3   2009    14      9       31    24
"))

DF
```


<br>

This data is considered wide since the *<u>time</u>* variable (represented as quarters) is structured such that each quarter represents a variable. To re-structure the time component as an individual variable, we can *gather* each quarter within one column variable and also *gather* the values associated with each quarter in a second column variable.

```{r, echo=TRUE, message=FALSE, collapse=TRUE}
long_DF <- DF %>% gather(Quarter, Revenue, Qtr.1:Qtr.4)
head(long_DF, 24)  # note, for brevity, I only show the data for the first two years 
```


```{r, echo=TRUE, eval=FALSE}
These all produce the same results:
        DF %>% gather(Quarter, Revenue, Qtr.1:Qtr.4)
        DF %>% gather(Quarter, Revenue, -Group, -Year)
        DF %>% gather(Quarter, Revenue, 3:6)
        DF %>% gather(Quarter, Revenue, Qtr.1, Qtr.2, Qtr.3, Qtr.4)

Also note that if you do not supply arguments for na.rm or convert values then the defaults are used
```

<a href="#">Go to top</a>

<br>
<a id="separate"> </a>

### <u>separate( ) function</u>: 
**Objective:** Splitting a single variable into two

**Description:** Many times a single column variable will capture multiple variables, or even parts of a variable you just don't care about.  Some examples include:

```{r, echo=FALSE, message=FALSE}
example_sep <- read.csv("/Volumes/Memorex/R Tutorial/Workbook1.csv")

example_sep
```

<br>

In each of these cases, our objective may be to *separate* characters within the variable string. This can be accomplished using the `separate()` function which turns a single character column into multiple columns.

**Complement to:** <a href="#unite">`unite()`</a>

```{r, echo=TRUE, eval=FALSE}
Function:       separate(data, col, into, sep = " ", remove = TRUE, convert = FALSE)
Same as:        data %>% separate(col, into, sep = " ", remove = TRUE, convert = FALSE)

Arguments:
        data:           data frame
        col:            column name representing current variable
        into:           names of variables representing new variables
        sep:            how to separate current variable (char, num, or symbol)
        remove:         if TRUE, remove input column from output data frame
        convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or 
                        factor as appropriate
```


**Example**

We can go back to our **long_DF** dataframe we created above in which way may desire to clean up or separate the *Quarter* variable.
```{r, echo=FALSE, message=FALSE}
head(long_DF)
```

<br>

By applying the `separate()` function we get the following:
```{r, echo=TRUE, message=FALSE, collapse=TRUE}
separate_DF <- long_DF %>% separate(Quarter, c("Time_Interval", "Interval_ID"))
head(separate_DF, 10)
```

```{r, echo=TRUE, eval=FALSE}
These produce the same results:
        long_DF %>% separate(Quarter, c("Time_Interval", "Interval_ID"))
        long_DF %>% separate(Quarter, c("Time_Interval", "Interval_ID"), sep = "\\.")
```

<a href="#">Go to top</a>

<br>
<a id="unite"> </a>

### <u>unite( ) function</u>: 
**Objective:** Merging two variables into one

**Description:** There may be a time in which we would like to combine the values of two variables.  The `unite()` function is a convenience function to paste together multiple variable values into one.  In essence, it combines two variables of a single observation into one variable.

**Complement to:**  <a href="#separate">`separate()`</a>

```{r, echo=TRUE, eval=FALSE}
Function:       unite(data, col, ..., sep = " ", remove = TRUE)
Same as:        data %>% unite(col, ..., sep = " ", remove = TRUE)

Arguments:
        data:           data frame
        col:            column name of new "merged" column
        ...:            names of columns to merge
        sep:            separator to use between merged values
        remove:         if TRUE, remove input column from output data frame
```

<br>

**Example**

Using the **separate_DF** dataframe we created above, we can re-unite the *Time_Interval* and *Interval_ID* variables we created and re-create the original *Quarter* variable we had in the **long_DF** dataframe.

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
unite_DF <- separate_DF %>% unite(Quarter, Time_Interval, Interval_ID, sep = ".")
head(unite_DF, 10)
```

```{r, echo=TRUE, eval=FALSE}
These produce the same results:
        separate_DF %>% unite(Quarter, Time_Interval, Interval_ID, sep = "_")
        separate_DF %>% unite(Quarter, Time_Interval, Interval_ID)

If no spearator is identified, "_" will automatically be used
```

<a href="#">Go to top</a>

<br>
<a id="spread"> </a>

### <u>spread( ) function</u>: 

**Objective:** Reshaping long format to wide format

**Description:** There are times when we are required to turn long formatted data into wide formatted data.  The `spread()` function spreads a key-value pair across multiple columns.

**Complement to:** <a href="#gather">`gather()`</a>

```{r, echo=TRUE, eval=FALSE}
Function:       spread(data, key, value, fill = NA, convert = FALSE)
Same as:        data %>% spread(key, value, fill = NA, convert = FALSE)

Arguments:
        data:           data frame
        key:            column values to convert to multiple columns
        value:          single column values to convert to multiple columns' values 
        fill:           If there isn't a value for every combination of the other variables and the key 
                        column, this value will be substituted
        convert:        if TRUE will automatically convert values to logical, integer, numeric, complex or 
                        factor as appropriate
```

**Example**

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
wide_DF <- unite_DF %>% spread(Quarter, Revenue)
head(wide_DF, 24)
```

<a href="#">Go to top</a>

<center><hr width="50%"></center>

<br>

## <u>dplyr Operations</u>

There are seven fundamental functions of data transformation:

* <a href="#select">`select()`</a> selecting variables
* <a href="#filter">`filter()`</a> provides basic filtering capabilities
* <a href="#group">`group_by()`</a> groups data by categorical levels
* <a href="#summarise">`summarise()`</a> summarise data by functions of choice
* <a href="#arrange">`arrange()`</a> ordering data
* <a href="#join">`join()`</a> joining separate dataframes
* <a href="#mutate">`mutate()`</a> create new variables

<br>

For these examples we'll use the following [census data](http://www.census.gov/en.html) which includes the K-12 public school expenditures by state.  This dataframe currently is 50x16 and includes expenditure data for 14 unique years.

Left half of data:
```{r, echo=FALSE, message=FALSE, cache=TRUE}
library(stringr)

expenditures <- read.csv("~/Desktop/Personal/Education & Training/Personal Analyses/Cost per H.S. Graduate/expenditures.csv")

# Format expenditure data
expenditures$State <- str_replace_all(expenditures$State, "[^[:alnum:]]", " ")
expenditures$State <- str_trim(expenditures$State)

expenditures <- expenditures %>% filter(State != "United States")

Regions <- read.csv("~/Desktop/Personal/Education & Training/AFIT/Dissertation/Paper #2/Data Analysis/Census_Bureau_State_Regions.csv")

expenditures <- expenditures %>% 
        left_join(Regions) %>% 
        select(Division, State, X1980:X2011) %>%
        filter(Division != "NA")

head(expenditures[,1:8])
```

Right half of data:
```{r, echo=FALSE, message=FALSE, cache=TRUE}
head(expenditures[,9:16])
```

<a href="#">Go to top</a>

<br>
<a id="select"> </a>

### <u>select( ) function</u>: 
**Objective:** Reduce dataframe size to only desired variables for current task

**Description:** When working with a sizable dataframe, often we desire to only assess specific variables.  The `select()` function allows you to select and/or rename variables.

```{r, echo=TRUE, eval=FALSE}
Function:       select(data, ...)
Same as:        data %>% select(...)

Arguments:
        data:           data frame
        ...:            call variables by name or by function

Special functions:
        starts_with(x, ignore.case = TRUE): names starts with x
        ends_with(x, ignore.case = TRUE):   names ends in x
        contains(x, ignore.case = TRUE):    selects all variables whose name contains x
        matches(x, ignore.case = TRUE):     selects all variables whose name matches the regular expression x
```

**Example**
Let's say our goal is to only assess the 5 most recent years worth of expenditure data.  Applying the `select()` function we can *select* only the variables of concern. 

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp <- expenditures %>% select(Division, State, X2007:X2011)
head(sub.exp)  # for brevity only display first 6 rows
```

<br>

We can also apply some of the special functions within `select()`. For instance we can select all variables that start with 'X':

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
head(expenditures %>% select(starts_with("X")))
```


```{r, echo=TRUE, eval=FALSE}
You can also de-select variables by using "-" prior to name or function.  The following produces the inverse of functions above
        expenditures %>% select(-X1980:-X2006)
        expenditures %>% select(-starts_with("X"))
```

<a href="#">Go to top</a>

<br>
<a id="filter"> </a>

### <u>filter( ) function</u>: 
**Objective:** Reduce rows/observations with matching conditions

**Description:** Filtering data is a common task to identify/select observations in which a particular variable matches a specific value/condition. The `filter()` function provides this capability. 

```{r, echo=TRUE, eval=FALSE}
Function:       filter(data, ...)
Same as:        data %>% filter(...)

Arguments:
        data:           data frame
        ...:            conditions to be met
```

<br>

**Examples**

Continuing with our **sub.exp** dataframe which includes only the recent 5 years worth of expenditures, we can filter by *Division*:
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>% filter(Division == 3)
```

<br>

We can apply multiple logic rules in the `filter()` function such as:
```{r, echo=TRUE, message=FALSE, cache=TRUE, eval=FALSE}
<   Less than                    !=      Not equal to
>   Greater than                 %in%    Group membership
==  Equal to                     is.na   is NA
<=  Less than or equal to        !is.na  is not NA
>=  Greater than or equal to     &,|,!   Boolean operators
```

<br>

For instance, we can filter for Division 3 and expenditures in 2011 that were greater than $10B.  This results in Indiana, which is in Division 3, being excluded since its expenditures were < $10B *(FYI - the raw census data are reported in units of $1,000)*.

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>% filter(Division == 3, X2011 > 10000000)  # Raw census data are in units of $1,000
```

<a href="#">Go to top</a>

<br>
<a id="group"> </a>

### <u>group_by( ) function</u>: 
**Objective:** Group data by categorical variables

**Description:** Often, observations are nested within groups or categories and our goals is to perform statistical analysis both at the observation level and also at the group level.  The `group_by()` function allows us to create these categorical groupings.

```{r, echo=TRUE, eval=FALSE}
Function:       group_by(data, ...)
Same as:        data %>% group_by(...)

Arguments:
        data:           data frame
        ...:            variables to group_by

*Use ungroup(x) to remove groups
```

<br>

**Example**
The `group_by()` function is a *silent* function in which no observable manipulation of the data is performed as a result of applying the function.  Rather, the only change you'll notice is, if you print the dataframe you will notice underneath the *Source* information and prior to the actual dataframe, an indicator of what variable the data is grouped by will be provided. The real magic of the `group_by()` function comes when we perform summary statistics which we will cover shortly. 

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
group.exp <- sub.exp %>% group_by(Division)

head(group.exp)
```

<a href="#">Go to top</a>

<br>
<a id="summarise"> </a>

### <u>summarise( ) function</u>:
**Objective:** Perform summary statistics on variables

**Description:** Obviously the goal of all this data *wrangling* is to be able to perform statistical analysis on our data.  The `summarise()` function allows us to perform the majority of the initial summary statistics when performing exploratory data analysis.

```{r, echo=TRUE, eval=FALSE}
Function:       summarise(data, ...)
Same as:        data %>% summarise(...)

Arguments:
        data:           data frame
        ...:            Name-value pairs of summary functions like min(), mean(), max() etc.

*Developer is from New Zealand...can use "summarise(x)" or "summarize(x)"
```

<br>

**Examples**

Lets get the mean expenditure value across all states in 2011
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>% summarise(Mean_2011 = mean(X2011))
```


Not too bad, lets get some more summary stats
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>% summarise(Min = min(X2011, na.rm=TRUE),
                     Median = median(X2011, na.rm=TRUE),
                     Mean = mean(X2011, na.rm=TRUE),
                     Var = var(X2011, na.rm=TRUE),
                     SD = sd(X2011, na.rm=TRUE),
                     Max = max(X2011, na.rm=TRUE),
                     N = n())
```


This information is useful, but being able to compare summary statistics at multiple levels is when you really start to gather some insights.  This is where the `group_by()` function comes in.  First, let's group by *Division* and see how the different regions compared in by 2010 and 2011.
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>%
        group_by(Division)%>% 
        summarise(Mean_2010 = mean(X2010, na.rm=TRUE),
                  Mean_2011 = mean(X2011, na.rm=TRUE))
```


Now we're starting to see some differences pop out.  How about we compare states within a Division?  We can start to apply multiple functions we've learned so far to get the 5 year average for each state within Division 3.
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>%
        gather(Year, Expenditure, X2007:X2011) %>%   # this turns our wide data to a long format
        filter(Division == 3) %>%                    # we only want to compare states within Division 3
        group_by(State) %>%                          # we want to summarize data at the state level
        summarise(Mean = mean(Expenditure),
                  SD = sd(Expenditure))
```

<a href="#">Go to top</a>

<br>
<a id="arrange"> </a>

### <u>arrange( ) function</u>: 
**Objective:** Order variable values

**Description:**  Often, we desire to view observations in rank order for a particular variable(s). The `arrange()` function allows us to order data by variables in accending or descending order.

```{r, echo=TRUE, eval=FALSE}
Function:       arrange(data, ...)
Same as:        data %>% arrange(...)

Arguments:
        data:           data frame
        ...:            Variable(s) to order

*use desc(x) to sort variable in descending order
```

<br>

**Examples**

For instance, in the summarise example we compared the the mean expenditures for each division. We could apply the `arrange()` function at the end to order the divisions from lowest to highest expenditure for 2011.  This makes it easier to see the significant differences between Divisions 8,4,1 & 6 as compared to Divisions 5,7,9,3 & 2.

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>%
        group_by(Division)%>% 
        summarise(Mean_2010 = mean(X2010, na.rm=TRUE),
                  Mean_2011 = mean(X2011, na.rm=TRUE)) %>%
        arrange(Mean_2011)
```


We can also apply an *descending* argument to rank-order from highest to lowest.  The following shows the same data but in descending order by applying `desc()` within the `arrange()` function.
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
sub.exp %>%
        group_by(Division)%>% 
        summarise(Mean_2010 = mean(X2010, na.rm=TRUE),
                  Mean_2011 = mean(X2011, na.rm=TRUE)) %>%
        arrange(desc(Mean_2011))
```


<a href="#">Go to top</a>

<br>
<a id="join"> </a>

### <u>join( ) functions</u>: 
**Objective:** Join two datasets together

**Description:** Often we have separate dataframes that can have common and differing variables for similar observations and we wish to *join* these dataframes together.  The multiple `xxx_join()` functions provide multiple ways to join dataframes.

```{r, echo=TRUE, eval=FALSE}
Description:    Join two datasets

Function:       
                inner_join(x, y, by = NULL)
                left_join(x, y, by = NULL)
                semi_join(x, y, by = NULL)
                anti_join(x, y, by = NULL)

Arguments:
        x,y:           data frames to join
        by:            a character vector of variables to join by. If NULL, the default, join will do a natural join, using all 
                        variables with common names across the two tables.
```

<br>

**Example**

Our public education expenditure data represents then-year dollars.  To make any accurate assessments of longitudinal trends and comparison we need to adjust for inflation.  I have the following dataframe which provides inflation adjustment factors for base-year 2012 dollars *(obviously I should use 2014 values but I had these easily accessable and it only serves for illustrative purposes)*.
```{r, echo=FALSE, message=FALSE, cache=TRUE}
inflation <- read.csv("~/Desktop/Personal/Education & Training/Personal Analyses/Cost per H.S. Graduate/inflation.csv")

tail(inflation)
```


To join to my expenditure data I obviously need to get my expenditure data in the proper form that allows my to join these two dataframes.  I can apply the following functions to accomplish this:
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}
long.exp <- sub.exp %>%
        gather(Year, Expenditure, X2007:X2011) %>%         # turn to long format
        separate(Year, into=c("x", "Year"), sep="X") %>%   # separate "X" from year value
        select(-x)                                         # remove "x" column

long.exp$Year <- as.numeric(long.exp$ Year)  # convert from character to numeric

head(long.exp)
```


I can now apply the `left_join()` function to join the inflation data to the expenditure data.  This aligns the data in both dataframes by the *Year* variable and then joins the remaining inflation data to the expenditure dataframe as new variables.
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}

join.exp <- long.exp %>% left_join(inflation)

head(join.exp)
```


To illustrate the other joining methods we can use these two simple dateframes:

```{r, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}
x <- data.frame(
  name = c("John", "Paul", "George", "Ringo", "Stuart", "Pete"),
  instrument = c("guitar", "bass", "guitar", "drums", "bass", "drums"))

y <- data.frame(
    name = c("John", "Paul", "George", "Ringo", "Brian"),
    band = c("TRUE", "TRUE", "TRUE",  "TRUE", "FALSE"))
```

Dataframe "x":
```{r, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}
x
```

Dataframe "y":
```{r, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}
y
```

<br>

`inner_join()`: Include only rows in both x and y that have a matching value
```{r, echo=TRUE, message=FALSE, cache=TRUE, warning=FALSE, collapse=TRUE}
inner_join(x,y)
```

<br>

`left_join()`: Include all of x, and matching rows of y
```{r, echo=TRUE, message=FALSE, cache=TRUE, warning=FALSE, collapse=TRUE}
left_join(x,y)
```

<br>

`semi_join()`: Include rows of x that match y but only keep the columns from x
```{r, echo=TRUE, message=FALSE, cache=TRUE, warning=FALSE, collapse=TRUE}
semi_join(x,y)
```

<br>

`anti_join()`: Opposite of semi_join
```{r, echo=TRUE, message=FALSE, cache=TRUE, warning=FALSE, collapse=TRUE}
anti_join(x,y)
```


<a href="#">Go to top</a>

<br>
<a id="mutate"> </a>

### <u>mutate( ) function</u>: 
**Objective:** Creates new variables

**Description:** Often we want to create a new variable that is a function of the current variables in our dataframe or even just add a new variable.  The `mutate()` function allows us to add new variables while preserving the existing variables.

```{r, echo=TRUE, eval=FALSE}
Function:       
                mutate(data, ...)
Same as:        data %>% mutate(...)                

Arguments:
        data:           data frame
        ...:            Expression(s)
```

<br>

**Examples**

If we go back to our previous **join.exp** dataframe, remember that we joined inflation rates to our non-inflation adjusted expenditures for public schools.  The dataframe looks like:

```{r, echo=FALSE, message=FALSE, cache=TRUE}
head(join.exp)
```


If we wanted to adjust our annual expenditures for inflation we can use `mutate()` to create a new inflation adjusted cost variable which we'll name *Adj_Exp*:
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}

inflation_adj <- join.exp %>% mutate(Adj_Exp = Expenditure/Inflation)

head(inflation_adj)
```


Lets say we wanted to create a variable that rank-orders state-level expenditures (inflation adjusted) for the year 2010 from the highest level of expenditures to the lowest.  
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}

rank_exp <- inflation_adj %>% 
        filter(Year == 2010) %>%
        arrange(desc(Adj_Exp)) %>%
        mutate(Rank = 1:length(Adj_Exp))

head(rank_exp)
```


If you wanted to assess the percent change in cost for a particular state you can use the `lag()` function within the `mutate()` function:
```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}

inflation_adj %>%
        filter(State == "Ohio") %>%
        mutate(Perc_Chg = (Adj_Exp-lag(Adj_Exp))/lag(Adj_Exp))

```


You could also look at what percent of all US expenditures each state made up in 2011.  In this case we use `mutate()` to take each state's inflation adjusted expenditure and divide by the sum of the entire inflation adjusted expenditure column.  We also apply a second function within `mutate()` that provides the cummalative percent in rank-order.  This shows that in 2011, the top 8 states with the highest expenditures represented over 50% of the total U.S. expenditures in K-12 public schools.  *(I remove the non-inflation adjusted Expenditure, Annual & Inflation columns so that the columns don't wrap on the screen view)*

```{r, echo=TRUE, message=FALSE, cache=TRUE, collapse=TRUE}

perc.of.whole <- inflation_adj %>%
        filter(Year == 2011) %>%
        arrange(desc(Adj_Exp)) %>%
        mutate(Perc_of_Total = Adj_Exp/sum(Adj_Exp),
               Cum_Perc = cumsum(Perc_of_Total)) %>%
        select(-Expenditure, -Annual, -Inflation)
        
head(perc.of.whole, 8)
```

<a href="#">Go to top</a>

<br>

<center><hr width="50%"></center>

<br>

## <u>Additional Resources</u>
This tutorial simply touches on the basics that these two packages can do. There are several other resources you can check out to learn more.  In addition, much of what I have learned and, therefore, much of the content in this tutorial is simply a modified regurgitation of the wonderful resources provided by [R Studio](http://www.rstudio.com/), [Hadley Wickham](https://twitter.com/hadleywickham), and [Garrett Grolemund](https://twitter.com/StatGarrett).

- R Studio's [Data wrangling with R and RStudio webinar](http://www.rstudio.com/resources/webinars/)
- R Studio's [Data wrangling GitHub repository](https://github.com/rstudio/webinars/blob/master/2015-01/wrangling-webinar.pdf)
- R Studio's [Data wrangling cheat sheet](http://www.rstudio.com/resources/cheatsheets/)
- Hadley Wickham’s dplyr tutorial at useR! 2014, [Part 1](http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/)
- Hadley Wickham’s dplyr tutorial at useR! 2014, [Part 2](http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-2/)
- Hadley Wickham's paper on [Tidy Data](http://vita.had.co.nz/papers/tidy-data.html)

<br>

<a href="#">Go to top</a>

<br>

*<font size="2">Special thanks to Tom Filloon and Jason Freels for providing constructive comments while developing this tutorial.</font>*
